# Results

In this section, we will use the data to try to find insights into the questions posed in the introduction.

```{r}
suppressWarnings(suppressMessages({
library(ggplot2)
library(dplyr)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tidytext))
library(ggrepel)
library(tidyr)
library(parcoords)
library(htmltools)
library(ggridges)
options(ggplot2.discrete.fill.warning = FALSE)
}))
```

```{r}
source("scripts/pseoe_preprocessing.R")
pseoe <- read_csv("data/pseoe_all.csv", show_col_types = FALSE)
degree_mapping <- read.csv("data/label_degree_level.csv")
cipcode_mapping <- read.csv("data/label_cipcode.csv")
institution_mapping <- read.csv("data/label_institution.csv")


pseoe <- preprocess_pseoe(
  pseoe = pseoe,
  degree_mapping = degree_mapping,
  cipcode_mapping = cipcode_mapping,
  institution_mapping = institution_mapping
)
#cat("Rows:", nrow(pseoe), "\nColumns:", ncol(pseoe), "\n")
```

## What are the top-earning fields of study within each degree type and their growth potential?

This analysis explores the earnings outcomes for the top-earning fields of each degree type. This is to highlight trends across early, mid, and long-term career stages (Year 1, Year 5, Year 10). Intuition tells us that more advanced degrees could yield better outcomes, but field-specific patterns could help influence a prospective college student's choice of major.

```{r, fig.width=10, fig.height=12}
#| warning: false
#| message: false

top_n <-5

plot_df <-pseoe |>
  group_by(degree_level, degree_label, cipcode_label) |>
  summarize(
    median_y1_earnings = mean(y1_p50_earnings, na.rm = TRUE),
    median_y5_earnings =mean(y5_p50_earnings, na.rm = TRUE),
    median_y10_earnings= mean(y10_p50_earnings, na.rm = TRUE),
    .groups="drop"
  ) |>
  filter(
    !is.na(median_y1_earnings) | 
    !is.na(median_y5_earnings)| 
    !is.na(median_y10_earnings)
  ) |>
  arrange(degree_level, desc(median_y10_earnings)) |>
  group_by(degree_label) |>
  slice_head(n =top_n) |>
  mutate(
    cipcode_label=reorder_within(
      cipcode_label,-median_y10_earnings, degree_label
    )
  ) |>
  pivot_longer(
    cols = starts_with("median"), 
    names_to="year", values_to ="median_earnings"
  ) |>
  mutate(
    year=factor(
      year, levels=c(
        "median_y1_earnings","median_y5_earnings", "median_y10_earnings"
      ),
      labels=c("Year 1", "Year 5", "Year 10")
    ),
    degree_label =factor(
      degree_label, levels=unique(degree_label[order(degree_level)])
    )
  )

suppressWarnings(
  ggplot(plot_df, aes(x= cipcode_label, y=median_earnings, fill =year)) +
    geom_bar(stat ="identity", position="dodge",na.rm = TRUE) +
    facet_wrap(~degree_label, scales="free_x", nrow =2, drop =TRUE) + 
    scale_x_reordered() + 
    scale_y_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
    theme_minimal(base_size =14) + 
    labs(
      title =paste("Top", top_n, "Average Median Earnings by Field of Study and Degree Type") ,
      x= "Field of Study",
      y="Average Median Earnings (2021 US Dollars)",
      fill ="Earnings Year"
    ) +
    theme(
      axis.text.x = element_text(angle =55, hjust =1, size=10),
      axis.title.y=element_text(margin = margin(r = 10)),
      strip.text = element_text(size = 12, face = "bold", margin = margin(b=5)),
      panel.grid.major.x= element_blank(),
      plot.margin = margin(t = 20, r= 10, b = 20, l = 20),
      legend.position="top", legend.direction="horizontal"
    )
)
```

The plot displays the top 5 earning fields of study for each degree type. While the intuition that higher degrees yield higher income holds true in general, there are notable exceptions.

Fields related to science and technology consistently dominate top-earning outcomes across all degree levels. This likely reflects the consistent demand and high value placed on general analytical skills that we often see in the labor market. However, healthcare professionals appear to significantly increase their earning potential with advanced degrees, while engineering fields tend to show steady earning growth across all degree levels.

Also, certain fields stand out for disproportionately high earnings at advanced degree levels. For example, graduates with Health Profession Master’s degrees often earn more in Year 1 than most Doctoral - Research graduates make even 10 years post-graduation. This might reflect the immediate applicability and demand for healthcare expertise, like possibly nursing.

So, in general, prospective students can consider that STEM fields lead to top-earning salaries with appreciable growth.

## How do earnings compare across states and career progression?

For many students, location plays a crucial role in choosing a college. While they may not plan to settle in the state where they attend college, it’s worth considering the earnings potential for graduates from that state - particularly when evaluating opportunities for internships or entry-level positions.

```{r, fig.width=10, fig.height=12}
#| warning: false
#| message: false

plot_df <- pseoe |>
  filter(!institution_state %in% c("Others", "US")) |>
  group_by(institution_state) |>
  summarize(
    median_y1_earnings = mean(y1_p50_earnings, na.rm = TRUE),
    median_y5_earnings = mean(y5_p50_earnings, na.rm = TRUE),
    median_y10_earnings = mean(y10_p50_earnings, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(
    !is.na(median_y1_earnings) | 
    !is.na(median_y5_earnings) | 
    !is.na(median_y10_earnings)
  ) |>
  mutate(
    sort_value = ifelse(is.na(median_y10_earnings), median_y5_earnings, median_y10_earnings),
    institution_state = reorder(institution_state, sort_value)
  ) |>
  pivot_longer(
    cols = starts_with("median"),
    names_to = "year",
    values_to = "median_earnings"
  ) |>
  mutate(
    year = factor(
      year, levels = c(
        "median_y1_earnings", "median_y5_earnings", "median_y10_earnings"
      ),
      labels = c("Year 1", "Year 5", "Year 10")
    )
  )

suppressWarnings(suppressMessages({suppressWarnings(
  ggplot(plot_df, aes(x = median_earnings, y = institution_state, color = year)) +
    geom_point(size = 3) + #, position = position_dodge(width = 0.5)
    scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = "."))+
    theme_minimal(base_size = 14) +
    labs(
      title = "Average Median Earnings by State",
      x = "Average Median Earnings (2021 US Dollars)",
      y = "State",
      color = "Earnings Year"
    ) +
    theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal"
    )
)
}))
```
This plot shows income outcomes by state for the thirty states covered in our dataset. For state-level analysis, it is important to note that there is no distinction on where the student was employed - that is to say that a Maine graduate could have been employed in South Carolina, but is still counted for Maine, for example. Also, due to regional cover gaps pointed out in the data section, we don't have data for each state. Due to incomplete coverage on the state level as well, we also have to point out that the calculated outcomes might be not fully representative because of a possible bias towards top or lower-ranking institutions.

The plot does show significant variation in income levels between states, highlighting disparities in educational quality, job opportunities or possibly the dominant or prevalent industries in each state. States with hubs for higher paying industries like manufacturing, finance, or technology, such as Texas or Pennsylvania, tend to have higher earnings while rural or agricultural states, such as Oklahoma or Utah, remain lower. 

Also, notably, a large population state with tech and financial sector hubs like New York shows a relatively low average median income in Year 1 but a significantly higher value by Year 10. This may reflect the variability and range of salaries in the state, driven by its diverse opportunities and the large population accessing them.

We can also observe the effects of the sampling bias in our data. Michigan tops the list with high earning outcomes, but this can be explained by the fact that the University of Michigan, a top-ranked public university, is the only institution in the dataset for this state.  ([PSEOE Data Coverage](https://lehd.ces.census.gov/data/pseo_experimental.html)).

Overall, what this means for prospective students is that where they go to school can be a general indicator for their starting salaries and career progression. In general, they should weigh the quality of the school's education against the opportunities local to that school, as taking such opportunities may impact their long term career trajectory. 

## Is there a correlation between Year 1 income outcomes and graduation counts by degree level?

Similar to location in picking a school, immediate earnings outcomes are a key consideration for many students when choosing their degree. The question arises from the following - are students more likely to choose degree paths that lead to higher initial (Year 1) earnings? If that’s the case, we would expect to see a positive correlation between Year 1 earnings outcomes and the number of students pursuing those higher paying degree types.

```{r, fig.width=11, fig.height=12}
#| warning: false
#| message: false

plot_df <- pseoe |>
  group_by(degree_level, degree_label) |>
  summarize(
    avg_median_y1_earnings = mean(y1_p50_earnings, na.rm = TRUE),
    total_employment = sum(y1_grads_earn, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(
    !is.na(avg_median_y1_earnings),
    total_employment > 0
  ) |>
  arrange(degree_level)

suppressWarnings(
  ggplot(plot_df, aes(y = total_employment, x = avg_median_y1_earnings)) +
    suppressWarnings(geom_point(size = 4, color = "red")) +
    suppressWarnings(geom_smooth(method = "lm",se = FALSE,color = "blue",linetype = "dashed")) +
    geom_label_repel(aes(label = degree_label), 
                     size = 4, 
                     max.overlaps = Inf, 
                     box.padding = 0.5, 
                     point.padding = 0.5, 
                     segment.color = "grey50") +
    theme_minimal(base_size = 14) +
    scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
    scale_y_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
    labs(
      title = "Total Graduation Counts vs. Average Median Earnings by Degree Type (Year 1)",
      y = "Total Graduation Count",
      x = "Average Median Earnings (2021 US Dollars)"
    ) +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title.x = element_text(size = 14, margin = margin(t = 10)),
      axis.title.y = element_text(size = 14, margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20)
    )
)

```

This plot shows Year 1 average median earnings against total graduation counts, with the dashed blue line corresponding to the regression line. While one might assume that higher degree levels - and their associated higher income - would attract more graduates, the data shows a rather negative correlation. High-paying degrees like Doctoral - Prof. and Doctoral - Research have relatively low graduation counts, while degrees like Baccalaureate and Associates have much higher counts despite lower incomes.

Certificate programs (\<1 year and 1-2 years) cluster at the low end for both earnings and employment, reinforcing their role as niche credentials, potentially for those who are looking to switch careers. Post-Bacc Certificates have low graduation counts but show higher average median earnings than Masters, also possibly due to its specialized nature leading to demand in highly specific niches. Masters degrees seem to balance moderate incomes and employment by occupying the middle ground.

Furthermore, the seemingly negative correlation between median earnings and graduation count can also be explained by how the more advanced the degree is, the harder it is to obtain. This leads to fewer graduates, while the skills obtained in these degrees increases accessibility to high paying jobs, leading to higher median earnings.

Overall, this plot clearly suggests that factors beyond immediate income, such as accessibility, cost, and career pathways, may heavily influence degree level choices. The job market does not require highly advanced degrees for most positions, so it could be an economical choice for most to pursue employment over continuing education at a cost, despite having better long term earnings potential than a standard Baccalaureate degree.

## How does the total number of graduates compare to employed graduates by degree level, one year from graduation?

Ideally, if graduation rates were in sync with demands from the job market, the gap between total employed graduates and total overall graduates would be relatively small. Here we will explore this gap, broken down by degree level. This analysis can help highlight which degree levels align more closely with immediate job market opportunities and where students might face greater challenges in securing employment.

```{r, fig.width=10, fig.height=10}
#| warning: false
#| message: false

plot_df <- pseoe |>
  group_by(degree_label, degree_level) |>
  summarize(
    total_graduates = sum(y10_ipeds_count, na.rm = TRUE),
    total_employment = sum(y10_grads_earn, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(
    total_graduates > 0,
    total_employment >= 0
  ) |>
  arrange(degree_level) |>
  mutate(
    degree_label = factor(degree_label, levels = unique(degree_label[order(degree_level)]))
  )

plot_df_long <- plot_df |>
  pivot_longer(
    cols = c(total_employment, total_graduates),
    names_to = "status",
    values_to = "count"
  ) |>
  mutate(
    status = factor(status, levels = c("total_employment", "total_graduates"),
                    labels = c("Employed", "Graduates"))
  )

suppressWarnings(
  ggplot(plot_df_long, aes(x = count, y = degree_label, color = status)) +
    geom_segment(data = plot_df, aes(x = total_employment, xend = total_graduates, 
                                     y = degree_label, yend = degree_label), 
                 inherit.aes = FALSE, color = "grey50", size = 1) +
    geom_point(size = 4) +
    scale_color_manual(values = c("red", "blue")) +
    theme_minimal(base_size = 14) +
    scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
    labs(
      title = "Graduate vs. Employment Counts by Degree Level (Year 1)",
      x = "Count of Students",
      y = "Degree Level",
      color = "Employment Status"
    ) +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title.x = element_text(size = 14, margin = margin(t = 10)),
      axis.title.y = element_text(size = 14, margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal"
    )
)

```

We see that niche or specialized degrees, such as certificates and doctoral degrees, show higher employment rates, with relatively small gaps between graduates and employed individuals. In contrast, more general degrees like Associates and Baccalaureate degrees show noticeable gaps, indicating that a significant portion of graduates remain unemployed.  Notably though, Baccalaureate degrees have the largest gap of these three. One reason for this could be that a significant portion of graduates of Baccaluareate degrees often opt to still pursue a Masters immediately after graduation, which would be counted as unemployed in our analysis.

However, the overall discrepancy for the general degrees when compared to the specialized degrees could be due to the broader applicability and generality of these degrees, which may lead to greater variability in employment outcomes. Additionally, the larger graduate populations for these degrees might create more competition in the job market, further contributing to the gaps. The relatively small graduate numbers in niche fields may align better with job market demand, which would result in the higher observed employment rates.

A common example familiar to us is those with Baccalaureate degrees trying to break into the Data Science industry having a harder time than those with a Masters degree. So, prospective students should consider the competition they may face and the demand for their degree when choosing a degree level.

## How do employment outcomes and growth compare across the most popular fields?

For students exploring various fields of study and weighing their long-term growth potential, comparing employment outcomes across fields is essential. Visualizing and normalizing the data allows us to make these comparisons more meaningful by putting different metrics, like employment rates and earnings, on the same scale. This way we can evaluate how fields stack up relative to one another, regardless of the raw differences in magnitude.

```{r, fig.width=11, fig.height=12}
#| warning: false
#| message: false


top_n <- 20
plot_df <- pseoe |>
  group_by(cipcode_label) |>
  summarize(
    total_graduates = sum(y1_ipeds_count, na.rm = TRUE),
    employment_rate = sum(y1_grads_earn, na.rm = TRUE) / total_graduates,
    avg_median_y1_income = mean(y1_p50_earnings, na.rm = TRUE),
    avg_median_y5_income = mean(y5_p50_earnings, na.rm = TRUE),
    avg_median_y10_income = mean(y10_p50_earnings, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(total_graduates > 0) |>
  arrange(desc(total_graduates)) |>
  slice_head(n = top_n)

plot_df_normalized <- plot_df |>
  mutate(
    'Degree Field' = cipcode_label,
    'Total Graduates' = total_graduates,
    'Employment Rate' = scales::rescale(employment_rate),
    'Y1 Earnings' = scales::rescale(avg_median_y1_income),
    'Y5 Earnings' = scales::rescale(avg_median_y5_income),
    'Y10 Earnings' = scales::rescale(avg_median_y10_income)
  )|>
  select(-avg_median_y1_income, -avg_median_y5_income, -avg_median_y10_income, -total_graduates, -employment_rate, -cipcode_label)

title_text<-sprintf("Employability and Income Analysis for Top %d Historically Popular Fields of Study (Normalized)", top_n)
title <- tags$h3(title_text)

# https://edav.info/multidimensional_continuous.html#interactive-parallel-coordinate-plot
parcoords_plot <- parcoords(
  data = plot_df_normalized,
  rownames = FALSE,
  brushMode = "1D-axes",
  color = list(colorBy = "Degree Field", colorScale = "scaleOrdinal", colorScheme = "schemeCategory10"),
  alpha = 0.5,
  withD3 = TRUE,
  #width = 1000,
  #height = 600
)

#scrollable_fix <- htmltools::div(
#  style = "overflow-y: auto; height: 600px; border: 1px solid #ccc; padding: 10px;",
#  parcoords_plot
#)

htmltools::browsable(tagList(title, parcoords_plot))
#htmltools::browsable(tagList(title, scrollable_fix))

```

This interactive plot provides an analysis of employment outcomes, including normalized employment rates and income trajectories, for the most popular fields of study (by graduate volume). Selections can be made on the plot to trace the paths of individual or groups of fields.

Fields like Homeland Security / Law Enforcement, Business Management, and Registered Nursing show high employment rates, which make them attractive for those seeking immediate job security. However, their normalized median incomes decline over time relative to other fields which suggests potential income stagnation for graduates in these areas.

Fields like Biological and Biomedical Engineering start with lower relative employment rates and initial incomes but demonstrate steady and appreciable growth in earnings over time. This indicates strong potential for career growth.

Also, while its employment rate remains middling compared to the other majors, Engineering achieves the highest normalized income growth, making it an excellent choice for those prioritizing long-term earning potential over initial job placement.

Overall, these trends highlight how different fields of study align with various career priorities - whether it's immediate employment or sustained income growth over time. A prospective student can weigh their personal goals to select a major by these metrics.

## What is the growth potential for different degree levels?

We already saw (and know) that different degree levels clearly result in different earnings outcomes, one, five or ten years after graduation. Related to this, we could also look at the growth potential for different types of degrees. Here we quantify the growth potential as the relation between the earnings one year and ten years after graduation.

```{r,fig.width=10, fig.height=9}
#| warning: false
#| message: false

pseoe_y1andy10 <- pseoe |>
                  filter(status_y1_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
                  filter(status_y10_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) 

ggplot(pseoe_y1andy10, aes(x = y1_p50_earnings, y = y10_p50_earnings)) +
  geom_point(alpha=0.2) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
  facet_wrap(~degree_label) +
  labs(
    x = "Year 1 Median Earnings (2021 US Dollars)",
    y = "Year 10 Median Earnings (2021 Us Dollars)",
    title = "Year 1 Median Earnings and Year 10 Median Earnings by Degree Type",
  ) +
  theme_minimal(base_size=14)
```

This figure shows scatterplots for median earnings one year versus ten years after graduation, one for each type of degree. In these plots, each dot represents one cohort. The red dashed line corresponds to "zero growth potential", i.e. the median earnings one and ten years after graduation is the same. We see that for all degree types, almost all cohorts have higher median earnings ten years after graduation. However, the range of magnitudes of this difference seems to depend strongly on the type of degree. The strongest growth potential seems to occur for cohorts in a professional doctoral program, where the median earnings 10 years after graduation can grow up to almost \$400,000 from around \$60,000 one year after graduation. Furthermore, Bachelors programs also allow for significant growth potential, where earnings ten years after graduation often seem to be \$100,000 higher than one year after graduation.

Research doctoral and Masters programs seem to have smaller increases in earnings one to ten years after graduation than Bachelors programs, but graduates from these programs do however start with higher earnings one year after graduation. The different types of certificates and associates degrees also seem to have some growth potential, with associates degrees showing increases in earnings of up to around \$80,000.

## Is a Masters worth it?

Should I get a Masters? This is a difficult question faced by many students who obtained their undergraduate degree. Here we try to provide insights into this question by looking at the difference in earnings for Bachelors versus Masters graduates in the same field of study.

```{r, fig.width=11, fig.height=16}
#| warning: false
#| message: false

pseoe_popular_cipcodes_Bach_Masters <- pseoe |>
  filter(status_y1_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  group_by(cipcode) |>
  filter(all(c("05", "07") %in% degree_level)) |>
  ungroup() |>
  filter(degree_level == "05" | degree_level == "07") |>
  group_by(cipcode) |>
  summarize(total_grads = sum(y1_ipeds_count, na.rm = TRUE)) |>
  arrange(desc(total_grads)) |>
  slice_head(n = 20)

top_cipcodes <- unique(pseoe_popular_cipcodes_Bach_Masters$cipcode)

pseoe_earnings_top_cipcode_Bach_Masters <- pseoe |>
  filter(cipcode %in% top_cipcodes) |>
  filter(degree_level == "05" | degree_level == "07") |>
  group_by(cipcode_label, degree_label) |>
  summarize(avg_y1_p50_earnings = mean(y1_p50_earnings, na.rm = TRUE),
            .groups = "drop") |>
  mutate(
    cipcode_label = factor(
      cipcode_label,
      levels = pseoe |>
        filter(cipcode %in% top_cipcodes, degree_level == "07") |>
        group_by(cipcode_label) |>
        summarize(avg_y1_p50_earnings = mean(y1_p50_earnings, na.rm = TRUE), .groups = "drop") |>
        arrange(avg_y1_p50_earnings) |>
        pull(cipcode_label)
    )
  )

ggplot(pseoe_earnings_top_cipcode_Bach_Masters, aes(x=avg_y1_p50_earnings, y=cipcode_label, color=degree_label)) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = "."),
                     limits = c(0, 110000)) +
  theme_minimal(base_size=14) +
  labs(
    x = "Year 1 Average Median Earnings (2021 US Dollars)",
    y = "Field of study",
    title = "Average Median Earnings for Baccalaureate v.s. Masters Graduates",
    color = "Degree Level"
  ) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal"
    )


```

This plot shows the average median earnings, one year after graduation, for the twenty most popular fields of study for which both Bachelors and Masters programs are available. We see that for all of these fields, the average median earnings is always highest for the Masters program. The magnitude of the difference depends on the specific field, where we see a difference of almost \$30,000 for business and management degrees and of more than \$20,000 for computer and information degrees. For engineering degrees a Masters degree only increases the average median earnings from about \$63,000 to around \$75,000, which is a small relative increase of around 20%. For other degree's such as history, psychology and English language, the absolute difference is also around \$10,000 but still corresponds to a relative change of about 33%.


```{r, fig.width=11, fig.height=16}
#| warning: false
#| message: false

pseoe_popular_cipcodes_Bach_Masters <- pseoe |>
  filter(status_y5_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  group_by(cipcode) |>
  filter(all(c("05", "07") %in% degree_level)) |>
  ungroup() |>
  filter(degree_level == "05" | degree_level == "07") |>
  group_by(cipcode) |>
  summarize(total_grads = sum(y5_ipeds_count, na.rm = TRUE)) |>
  arrange(desc(total_grads)) |>
  slice_head(n = 20)

top_cipcodes <- unique(pseoe_popular_cipcodes_Bach_Masters$cipcode)


pseoe_earnings_top_cipcode_Bach_Masters <- pseoe |>
  filter(cipcode %in% top_cipcodes) |>
  filter(degree_level == "05" | degree_level == "07") |>
  group_by(cipcode_label, degree_label) |>
  summarize(avg_y5_p50_earnings = mean(y5_p50_earnings, na.rm = TRUE), .groups = "drop") |>
  mutate(
    cipcode_label = factor(
      cipcode_label,
      levels = pseoe |>
        filter(cipcode %in% top_cipcodes, degree_level == "07") |>
        group_by(cipcode_label) |>
        summarize(avg_y5_p50_earnings = mean(y5_p50_earnings, na.rm = TRUE), .groups = "drop") |>
        arrange(avg_y5_p50_earnings) |>
        pull(cipcode_label)
    )
  )
ggplot(pseoe_earnings_top_cipcode_Bach_Masters, aes(x=avg_y5_p50_earnings, y=cipcode_label, color=degree_label)) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = "."),
                     limits = c(0, 110000)) +
  theme_minimal(base_size=14) +
  labs(
    x = "Year 5 Average Median Earnings (2021 US Dollars)",
    y = "Field of study",
    title = "Average Median Earnings for Baccalaureate v.s. Masters Graduates",
    color = "Degree Level"
  ) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal"
    )

```

This plot then again shows the average median earnings for both Bachelors and Masters programs, but now five years after graduation. We see again that the Masters programs lead to higher average median earnings. However, when closely comparing the two graphs, we see that for lower earning fields such as English language and history, the gap between Masters and Bachelors graduates seems to have slightly tightened. For higher earning field such as Computer and Engineering however, the gap seems to have slightly widened.

These graphs seems to indicate that a Masters degree is indeed worth it, especially for higher earning fields. However, this analysis does not take into account the cost of a Masters degree, both from a financial perspective and a time perspective. Completing a Masters degree requires a significant financial investment and the devoted time can also be used to gain work experience and progress your career, which might perhaps lead to higher earnings. Furthermore, this analysis also does not take into account at which stage in their career graduates are. Masters students often already have several years of work experience before starting the program, which might automatically lead to higher earnings after completing the program.

## How important is the institution where you earned your degree?

We all know that institutions have significant differences in reputation and (at least perceived) quality of education. However, we might ask ourselves how this translates into measurable outcomes such as earnings after graduation.

```{r, fig.width=10, fig.height=16}
#| warning: false
#| message: false


# Identify top 20 degree-CIP code combinations by total graduates
pseoe_popular_degrees <- pseoe |>
  filter(cipcode != "00") |>
  filter(status_y1_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  filter(status_y1_ipeds_count %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  group_by(degree_label, cipcode_label) |>
  summarize(total_grads = sum(y1_ipeds_count, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_grads)) |>
  slice_head(n = 20) |>
  mutate(cipcode_degree_label = paste(degree_label, cipcode_label, sep = " - "))

# Calculate median earnings for ordering
pseoe_ridgeline_data <- pseoe |>
  filter(cipcode != "00") |>
  filter(status_y1_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  filter(status_y1_ipeds_count %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  mutate(cipcode_degree_label = paste(degree_label, cipcode_label, sep = " - ")) |>
  filter(cipcode_degree_label %in% pseoe_popular_degrees$cipcode_degree_label) |>
  group_by(cipcode_degree_label, institution) |>
  summarize(avg_y1_p50_earnings = mean(y1_p50_earnings, na.rm = TRUE), .groups = "drop") |>
  group_by(cipcode_degree_label) |>
  mutate(median_earnings = median(avg_y1_p50_earnings, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(median_earnings)) |>
  mutate(cipcode_degree_label = factor(cipcode_degree_label, levels = rev(unique(cipcode_degree_label))))


# Ridgeline plot
suppressMessages(suppressWarnings(ggplot(pseoe_ridgeline_data, aes(x = avg_y1_p50_earnings, y = cipcode_degree_label)) +
  geom_density_ridges(alpha = 0.8, scale = 1.5, fill = "cornflowerblue") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = "."),
                     limits = c(0, 150000)) +
  labs(
    x = "Year 1 Median Earnings (2021 US Dollars)",
    y = "Degree Level & Field",
    title = "Distribution of Year 1 Median Earnings over Institutions,\n by Degree Level & Field",
  ) +
  theme_minimal(base_size=14) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal",
      plot.title = element_text(hjust = 0.5)
    )))

```

This ridgeline plot shows how the median earnings one year after graduation are distributed over institutions offering these degrees. These distributions are shown for the top 20 most popular degrees in the dataset. We see that the shape of the distribution varies strongly depending on the specific degree. For example, for a Masters in Business and management or a Bachelors in Health Professions, the distribution is widely spread. This indicates that for these degrees, the institution does matter significantly, as different institutions correspond to strongly differing median earnings.

However, for other degrees such as a Bachelors in English Language or Psychology, and to a smaller extent a Bachelors in Engineering, the distribution is much more tightly spread. Therefore for these degrees, the awarding institution seems to have a smaller impact on earnings outcomes.

In general, the graph also seems to exhibit a trend where higher earning degrees tend to have a more widely spread distribution.

```{r, fig.width=10, fig.height=16}
#| warning: false
#| message: false

# Calculate median earnings for ordering
pseoe_ridgeline_data <- pseoe |>
  filter(cipcode != "00") |>
  filter(status_y1_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  filter(status_y1_ipeds_count %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
  mutate(cipcode_degree_label = paste(degree_label, cipcode_label, sep = " - ")) |>
  filter(cipcode_degree_label %in% pseoe_popular_degrees$cipcode_degree_label) |>
  group_by(cipcode_degree_label, institution) |>
  summarize(avg_y1_p75_earnings = mean(y1_p75_earnings, na.rm = TRUE), .groups = "drop") |>
  group_by(cipcode_degree_label) |>
  mutate(median_earnings = median(avg_y1_p75_earnings, na.rm = TRUE)) |>
  ungroup() |>
  arrange(desc(median_earnings)) |>
  mutate(cipcode_degree_label = factor(cipcode_degree_label, levels = rev(unique(cipcode_degree_label))))


# Ridgeline plot
suppressMessages(suppressWarnings(ggplot(pseoe_ridgeline_data, aes(x = avg_y1_p75_earnings, y = cipcode_degree_label)) +
  geom_density_ridges(alpha = 0.8, scale = 1.5, fill = "cornflowerblue") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = "."),
                     limits = c(0, 150000)) +
  labs(
    x = "Year 1 75th Percentile Earnings (2021 US Dollars)",
    y = "Degree Level & Field",
    title = "Distribution of Year 1 75th Percentile Earnings over Institutions,\n by Degree Level & Field",
  ) +
  theme_minimal(base_size=14) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal",
      plot.title = element_text(hjust = 0.5)
    )))


```

This ridgeline plot then shows the distribution of the 75th percentile of earnings one year after graduation over different institutions. We see similar trends as in the previous graph, where the same degrees as before seem to have the most widely spread distributions. However, in general we observe that for each degree the 75th percentile earnings are more widely spread than the median earnings. This indicates that the awarding institution seems to be more important when considering the possibility for high earnings outcomes.

## How have career prospects changed over time?

The job market is ever-changing. Degrees in fields that led to high paying jobs twenty years ago might not do so presently, while other degrees might have seen a big increase in earnings. In this section, we look at how average median earnings for different degrees have changed from 2001 to 2016, which correspond to the earliest and most recent graduation cohorts in the dataset.

In an attempt to eliminate any bias that is present in the dataset due to incomplete reporting, we performed matching of cohorts from 2001 to cohorts of 2016. Specifically, for each cohort that graduated in 2001 from a certain institution with a certain degree, there is exactly one cohort that graduated from the same institution with the same degree in 2016, and vice-versa. This makes sure that when we are comparing earnings outcomes for a certain degree in 2001 versus 2016, we are doing so for the same institutions awarding the degree.

```{r, fig.width=11, fig.height=10}
#| warning: false
#| message: false

idcols_inst <- c(
  "agg_level_pseo", "inst_level", "institution", "degree_level",
  "cip_level", "cipcode" 
)

pseoe_01and16 <- pseoe |> 
                      filter(grad_cohort == "2001" | grad_cohort == "2016") |>
                      filter(status_y1_earnings %in% c(1, 2, 4, 6, 7, 9, 10, 12)) |>
                      group_by(across(all_of(idcols_inst))) |>
                      filter(all(c("2001", "2016") %in% grad_cohort))

# assertion
num_cohorts_check <- pseoe_01and16 |>
  summarize(num_cohorts = n_distinct(grad_cohort), .groups = "drop") |>
  summarize(all_two = all(num_cohorts == 2))

stopifnot(num_cohorts_check$all_two)

pseoe_cipcode_change01to16 <- pseoe_01and16 |>
                              ungroup() |>
                              group_by(cipcode_label, degree_label, grad_cohort) |>
                              summarise(mean_med_earning = mean(y1_p50_earnings, na.rm = TRUE), .groups = "drop") |>
                              ungroup() |>
                              pivot_wider(names_from = grad_cohort, values_from = mean_med_earning, 
                                          names_prefix ="earnings_") |> 
                              mutate(change_med_earning = earnings_2016 - earnings_2001)

pseoe_cipcode_change01to16 <- pseoe_cipcode_change01to16 |>
  arrange(change_med_earning) |>
  select(cipcode_label, degree_label, change_med_earning) |>
  mutate(cipcode_degree_label = paste( degree_label, cipcode_label, sep = " - "))


ggplot(tail(pseoe_cipcode_change01to16, 15), aes(y = factor(cipcode_degree_label,
                                                                   levels = cipcode_degree_label[order(change_med_earning)]),
                                                        x = change_med_earning)) +
  geom_col(fill = "cornflowerblue") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
  labs(
    x = "Change in Average Year 1 Median Earnings,\n (2021 US Dollars)",
    y = "Degree Level & Field",
    title = "Degrees with Largest Increase in Average Year 1 Median Earnings,\n (2001 to 2016)",
  ) +
  theme_minimal(base_size=14) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal",
      plot.title = element_text(hjust = 0.5)
    )
```

This plot shows the 15 degrees with the highest increase in average median earnings one year after graduation from 2001 to 2016. We see that a research doctorate in Computer and Information has seen the biggest increase of \$30,000. Also other computer-related degrees such as a Bachelors in computer engineering, computer/information and computer science make the top 15, where these have seen an increase of around \$12,000. A Bachelors in statistics has seen the second biggest and a research doctorate in business and management the third biggest increase of around \$20,000 and \$15,000 respectively.

```{r, fig.width=11, fig.height=10}
#| warning: false
#| message: false

ggplot(pseoe_cipcode_change01to16[1:15, ], aes(y = factor(cipcode_degree_label,
                                                                   levels = cipcode_degree_label[order(change_med_earning)]),
                                                        x = change_med_earning)) +
  geom_col(fill = "cornflowerblue") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = ".")) +
  labs(
    x = "Change in Average Year 1 Median Earnings,\n (2021 US Dollars)",
    y = "Degree Level & Field",
    title = "Degrees with Largest Decrease in Average Year 1 Median Earnings,\n (2001 to 2016)",
  ) +
  theme_minimal(base_size=14) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal",
      plot.title = element_text(hjust = 0.5)
    )
```

This plot then shows the degrees with the highest decrease in average median earnings one year after graduation from 2001 to 2016. We see that according to this dataset, a Bachelors in Pharmacy and Pharmaceutical sciences has seen the most drastic decrease of around \$70,000. However, when we examined the data more closely, we saw that this decrease is calculated from only 3 cohorts in both 2001 and 2016, which might be a too small sample to make any hard conclusions. We do however also see a big decrease of around \$25,000 for a professional doctorate in Pharmacy and Pharmaceutical sciences, where this does correspond to a significant sample size of 21 cohorts.

We also see a big decrease of around \$30,000 for a research doctorate in Architecture and a decrease of around \$25,000 for an associates degree in Apparel and Textiles.

```{r, fig.width=11, fig.height=6}
#| warning: false
#| message: false

pharmacy_01and16_bac <- pseoe_01and16 |>
                    filter(cipcode_label == "Pharmacy, Pharmaceutical...", degree_label == "Baccalaureate")
stopifnot(nrow(pharmacy_01and16_bac) == 3+3)

pharmacy_01and16_doct <- pseoe_01and16 |>
                    filter(cipcode_label == "Pharmacy, Pharmaceutical...", degree_label == "Doctoral - Prof.")
stopifnot(nrow(pharmacy_01and16_doct) == 21+21)

pseoe_typedegree_change01to16 <- pseoe_01and16 |>
                              group_by(degree_label, grad_cohort) |>
                              summarise(mean_med_earning = mean(y1_p50_earnings, na.rm = TRUE), .groups = "drop") |>
                              ungroup() |>
                              pivot_wider(names_from = grad_cohort, values_from = mean_med_earning, 
                                          names_prefix ="earnings_") |> # Reshape data
                              mutate(change_med_earning = earnings_2016 - earnings_2001) # Compute change

pseoe_typedegree_change01to16 <- pseoe_typedegree_change01to16 |>
  arrange(change_med_earning) |>
  select(degree_label, change_med_earning)

ggplot(pseoe_typedegree_change01to16, aes(y = factor(degree_label,
                                                                   levels = degree_label[order(change_med_earning)]),
                                                        x = change_med_earning)) +
  geom_col(fill = "cornflowerblue") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",", decimal.mark = "."),
                     limits = c(-6000, 1000)) +
  labs(
    x = "Change in Average Year 1 Median Earnings,\n (2001 to 2016)",
    y = "Degree Level",
    title = "Change in Average Year 1 Median Earnings by Degree Level,\n (2021 US Dollars)",
  ) +
  theme_minimal(base_size=14) +
  theme(
      axis.text.y = element_text(size = 10),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(margin = margin(r = 10)),
      plot.margin = margin(t = 20, r = 10, b = 20, l = 20),
      legend.position = "top", legend.direction = "horizontal",
      plot.title = element_text(hjust = 0.5)
    )
```

This plot then shows the change in average median earnings one year after graduation for eight different degree levels. We see that only 2 degree levels - a professional doctorate and a one to two year certificate - have seen a small increase in median earnings of less then \$500. A Masters degree and a two to four year certificate have seen the biggest decreases of around \$1000 and \$5,500 respectively. Note also that due to the matching process, the earnings decrease for Masters programs for example, cannot be caused by the addition of new types of Masters programs that are not attractive in the job market. The decrease is only calculated for Masters for which cohorts graduated in both 2001 and 2016.
